Type theory in type theory with single substitutions

### 2024-09-15 types_2024_postproc/p.tex
stuff

\usepackage{tikz}
\usetikzlibrary{arrows}

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}

@@

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd
\end{alignedat}
\end{equation}
We use implicit arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly,
and $\app$ takes $i$, $A$ and $B$ implicitly. $f : X \cong Y : g$
denotes an isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with
$g\,(f\,x) = x$ and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$
and $\lam$ are second-order functions. We say that $\Pi$ binds a
$\Tm$-variable of type $A$ in its second $\Ty$-argument. Every binder
is represented by a second-order function, and as a result, we don't
need to talk about contexts and substitutions when describing the
theory: we re-use the binding features of the metatheory to describe
those of our object theory (this technique is also known as
higher-order abstract syntax \cite{DBLP:conf/lics/Hofmann99} or
logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
is no good notion of homomorphism between second-order models, thus
the semantics of a SOGAT is given by a first order generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts. The usual results from universal algebra (e.g.\ models
form a category with finite limits and initial object, existence of
free and co-free models) carry over to GATs
\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
theory of categories where there is a sort for objects and a sort for
morphisms double-indexed over objects.

@@

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item Minimisation.
\item For type theory with $\Pi$, $\U$, $\top$ and $\Sigma$, deriving
  a parallel model from a single substitution model.
\end{itemize}

@@

In the 1990s single substitution calculi for type theory were popular,
but usually in the extrinsic setting, see e.g.\ the thesis of
Altenkirch \cite{alti:phd93}. B-systems were introduced by Voevodsky
\cite{voevodsky2014bsystems} as an algebraic way of describing type
theories close to the notations using typing judgements. B-systems are
a generalised algebraic (intrinsic) presentation of type theory using
single substitutions. xxx

\begin{tikzpicture}
  \node (b0) at (0,0) {$\top\cong B_0$};
  \node (b1) at (2,0) {$B_1$};
  \node (b2) at (4,0) {$B_2$};
  \node (b3) at (6,0) {$\dots$};
  \node (c1) at (3,1) {$\tilde{B}_1$};
  \node (c2) at (5,1) {$\tilde{B}_1$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
\end{tikzpicture}

Using our
notations B-systems can be described as follows. B-frame. TODO

we use Sub, but we could have used telescopes

The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like in our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which makes it more apparent that
substitutions are simple lists of terms. CwFs are equivalent to
Ehrhard's calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

@@

Most formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than ordinary Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. QIITs are supported by the cubical
set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model \cite{kaposi-qiit-setoid}, thus
Cubical Agda \cite{DBLP:journals/jfp/VezzosiMA21} and
setoid/observational type theory \cite{setoid,DBLP:phd/hal/Pujet22} support
QIITs. For example, in Cubical Agda, set-truncated and
groupoid-truncated syntaxes of type theory have been formalised
\cite{cohtt}. Without QIITs, tricks such as shallow embedding the
syntax can be used to formalise metatheoretic results about type
theory such as gluing \cite{kaposi_et_al:LIPIcs:2019:10532}, and its
special cases canonicity and parametricity \cite{kaposi-shallow}.

At the level of abstraction of SOGATs, the difference between single
and parallel substitution calculi are not visible. However, the
metatheory of type theory can be still investigated in this abstract
setting, via the methods of Synthethic Tait Computability
\cite{DBLP:phd/us/Sterling22} or internal sconing
\cite{DBLP:conf/fscd/BocquetKS23}.

@@

Notation and formalisation techniques.

@@

Next we add dependent function space $\Pi$: the domain of a dependent
function is a type in some context $\Gamma$, and the codomain can also
refer to a variable in the domain, so we extend the context of the
codomain type with the type of the domain. For simplicity, both types
are at the same level.
\begin{alignat*}{10}
& \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i
\end{alignat*}
The nondependent function space $\blank\Ra\blank :
\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i$ is a special case of $\Pi$
and we define it as the abbreviation $A\Ra B := \Pi\,A\,(B[\p])$.

@@

With the introduction of $\blank\cdot\blank$, we need a new
substitution rule, but again it only makes sense with an extra
equation on types saying that first substituting the last variable and
then an arbitrary substitution is the same as first the lifted version
of the arbitrary substitution that does not touch the last variable,
and then substituting the last variable.
\begin{alignat*}{10}
& [\langle\rangle] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
\end{alignat*}
Following the introduction of the operation $\langle\blank\rangle$, we
need to explain how it acts on variables. Given a variable in the
middle of the context (a term $b[\p]$), substituting its last variable
simply returns $b$. Substituting the last variable $\q$ reads out the
term from $\langle\blank\rangle$. As usual, the rules only typecheck
if we have an equation for types.
\begin{alignat*}{10}
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a
\end{alignat*}
We don't have a separate sort of variables, so the rule
$[\p][\langle\rangle]$ holds not only for variables, but for arbitrary
terms. This is not an issue, as weakening a term at the end of its
context, and then substituting the newly introduced variable is the
same as not doing anything.

@@

Finally, we add the rules for universes. A universe is a type
containing codes for types, this is witnessed by $\El$ (elements) and
$\cd$ (code), which make an isomorphism between terms of type $\U\,i$
and types of level $i$. All three operations come with substitution
rules ($\El[]$ and $\cd[]$ are interderivable, so it would be enough
to assume one of the two, but we add both to completely align with the
algorithm \cite{DBLP:conf/fscd/KaposiX24} generating the rules from
the SOGAT (\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (n:\N)\ra\Ty\,\Gamma\,(1+n) \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A} \\
  & \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma])
\end{alignat*}

Show examples of using the calculus, and where we need to derive more
stuff.

no need for parallel substitutions, empty substitution, parallel
weakenings, telescopes, or combinations of these.

say that we assume the QIIT version of this.

\section{$\alpha$-normalisation}
\label{sec:alpha}

@@

\section{Isomorphism}
\label{sec:cwf}

@@

\section{Minimisation}

>>>>

\usepackage{tikz}
\usetikzlibrary{arrows}
\usepackage{proof}

\newcommand{\ra}{\rightarrow}
\newcommand{\Ra}{\Rightarrow}
\newcommand{\Set}{\mathsf{Set}}
\newcommand{\Ty}{\mathsf{Ty}}
\newcommand{\Tm}{\mathsf{Tm}}
\newcommand{\Con}{\mathsf{Con}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\p}{\mathsf{p}}
\newcommand{\q}{\mathsf{q}}
\newcommand{\ext}{\mathop{\triangleright}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\lam}{\mathsf{lam}}
\newcommand{\app}{\mathsf{app}}
\newcommand{\U}{\mathsf{U}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\cd}{\mathsf{c}}
\newcommand{\blank}{\mathord{\hspace{1pt}\text{--}\hspace{1pt}}} %from the book
\renewcommand{\tt}{\mathsf{tt}}
\newcommand{\fst}{\mathsf{fst}}
\newcommand{\snd}{\mathsf{snd}}
\newcommand{\Lift}{\mathsf{Lift}}
\newcommand{\mk}{\mathsf{mk}}
\newcommand{\un}{\mathsf{un}}

@@

Type theory is a language which can be described as a second-order
generalised algebraic theory (SOGAT
\cite{DBLP:conf/fscd/KaposiX24}). Some properties of this description:
\begin{enumerate}
\item[(i)] It is an intrinsic presentation
  \cite{DBLP:conf/csl/AltenkirchR99,DBLP:conf/popl/AltenkirchK16} as
  opposed to extrinsic \cite{abel2013normalization,theo}. That is, we
  only consider well-formed, well-scoped and well-typed abstract
  syntax trees, there are no meaningless terms.
\item[(ii)] Terms are quotiented by the conversion relation.
\item[(iii)] Every operation is stable under substitution.
\end{enumerate}
For example, type theory with $\Pi$ types and universes {\`a} la
Coquand \cite{coquandUniverse} is specified by the following SOGAT.
\begin{equation}\label{eq:tt}
\begin{alignedat}{10}
  & \Ty && : \N\ra\Set \\
  & \Tm && : \Ty\,i\ra\Set \\
  & \Pi && : (A:\Ty\,i)\ra(\Tm\,A\ra\Ty\,i)\ra\Ty\,i \\
  & \app && : \Tm\,(\Pi\,A\,B)\cong((a:\Tm\,A)\ra\,\Tm\,(B\,a)): \lam \\
  & \U && : (i:\N)\ra\Ty\,(1+i) \\
  & \El && : \Tm\,(\U\,i)\cong\Ty\,i : \cd \\
  & \Lift && : \Ty\,i\ra\Ty\,(1+i) \\
  & \un && : \Tm\,(\Lift\,A)\cong\Tm\,A : \mk
\end{alignedat}
\end{equation}
Because types are isomorphic to a term (witnessed by $\El$ and $\cd$),
we have to index types by their (universe) level, and types at lower
level are included at all upper levels by $\Lift$. We use implicit
arguments, e.g.\ $\Tm$ takes the $i:\N$ implicitly, and $\app$ takes
$i$, $A$ and $B$ implicitly. $f : X \cong Y : g$ denotes an
isomorphism, that is, $f : X\ra Y$, $g:Y\ra X$ with $g\,(f\,x) = x$
and $f\,(g\,y) = y$ for all $x$, $y$. Note that $\Pi$ and $\lam$ are
second-order functions. We say that $\Pi$ binds a $\Tm$-variable of
type $A$ in its second $\Ty$-argument. Every binder is represented by
a second-order function, and as a result, we don't need to talk about
contexts and substitutions when describing the theory: we re-use the
binding features of the metatheory to describe those of our object
theory (this technique is also known as higher-order abstract syntax
\cite{DBLP:conf/lics/Hofmann99} or logical framework
\cite{DBLP:journals/jacm/HarperHP93,10.1007/978-3-642-14203-1_2}). There
is no good notion of homomorphism between second-order models, thus
the semantics of a SOGAT is given by a first order generalised
algebraic theory (GAT \cite{DBLP:journals/apal/Cartmell86}). GATs are
generalisations of single-sorted algebraic theories (such as monoids,
groups) to multiple sorts where later sorts can be indexed over
previous sorts. The usual results from universal algebra (e.g.\ models
form a category with finite limits and initial object, existence of
free and co-free models) carry over to GATs
\cite{andras,DBLP:phd/hal/Moeneclaey22}. An example of a GAT is the
theory of categories where there is a sort for objects and a sort for
morphisms double-indexed over objects.

@@

To summarise, our contributions are the following:
\begin{itemize}
\item A new generalised algebraic presentation of type theory in the
  form of a minimalistic single substitution calculus. We show that
  there is no need for parallel substitutions, empty substitution,
  parallel weakenings, telescopes, or combinations of these to
  describe the syntax of type theory in an intrinsic way.
\item The $\alpha$-normalisation technique and deriving parallel
  substitutions from the single substitution syntax.
\item Minimisation.
\item For type theory with $\Pi$, $\U$, $\top$ and $\Sigma$, deriving
  a parallel model from a single substitution model.
\end{itemize}

@@

In the 1990s single substitution calculi for type theory were popular,
but usually in the extrinsic setting, see e.g.\ the thesis of
Altenkirch \cite{alti:phd93}.

B-systems were introduced by Voevodsky \cite{voevodsky2014bsystems} as
an algebraic way of describing type theories close to the notations
using typing judgements. B-systems are an intrinsic, but essentially
algebraic presentation of type theory using single
substitutions. B-systems correspond to our single substitution
calculus as essentially algebraic theories correpond to generalised
algebraic theories, or sets with a map into $I$ correspond to indexed
families over $I$ \cite[page
  221]{DBLP:journals/apal/Cartmell86}. Another difference is that we
have fewer and less general equations, resulting in the fact that we
have more models than B-systems. However in the syntax of our theory,
all the rules of B-systems are admissible. We describe the
relationship in more detail. B-systems are B-frames together with a
substitution, weakening and generic element operations. A B-frame is
given by sets $B_i$, $\tilde{B}_i$ and functions between them as shown
below \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}. \\
\begin{tikzpicture}
  \node (b0) at (0,0) {$\top$};
  \node (b1) at (1.5,0) {$B_1$};
  \node (b2) at (4,0) {$B_2$};
  \node (b3) at (7,0) {$\dots$};
  \node (c1) at (2.5,1) {$\tilde{B}_1$};
  \node (c2) at (5,1) {$\tilde{B}_1$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{ft}_0$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{ft}_1$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{ft}_2$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\partial_1$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\partial_2$} (b2);
\end{tikzpicture} \\
Using our notation, a B-frame contains the following data. \\
\begin{tikzpicture}
  \node (b0) at (0,0) {$\top$};
  \node (b1) at (1.5,0) {$\Ty\,\diamond$};
  \node (b2) at (4,0) {$(A:\Ty\,\diamond)\times\Ty\,(\diamond\ext A)$};
  \node (b3) at (7,0) {$\dots$};
  \node (c1) at (2.5,1) [text width=3cm] {$(A:\Ty\,\diamond)\times\Tm\,\diamond\,A$};
  \node (c2) at (5.7,1.7) [text width=4cm,align=center] {$\big((A:\Ty\,\diamond)\times(B:\Ty\,(\diamond\ext A))\big)\times\Tm\,(\diamond\ext A)\,B$};
  \draw[->,font=\scriptsize] (b1) edge node[below] {$\mathsf{fst}$} (b0);
  \draw[->,font=\scriptsize] (b2) edge node[below] {$\mathsf{fst}$} (b1);
  \draw[->,font=\scriptsize] (b3) edge node[below] {$\mathsf{fst}$} (b2);
  \draw[->,font=\scriptsize] (c1) edge node[above] {$\mathsf{fst}$} (b1);
  \draw[->,font=\scriptsize] (c2) edge node[above] {$\mathsf{fst}$} (b2);
\end{tikzpicture} \\
The substitution operation $\mathbb{S}$ for an element of
$x:\tilde{B}_{n+1}$ is a homomorphism of the slice B-frames
$\mathbb{B}/\partial(x) \ra \mathbb{B}/\mathsf{ft}(\partial(x))$. In
our notation, $x = (\Gamma, A, a)$ where $a : \Tm\,\Gamma\,A$, and the
homomorphism $\mathbb{S}$ corresponds to the following maps for the
bottom row of the diagram:
\begin{alignat*}{10}
  & \blank[\langle a\rangle] && : \Ty\,(\Gamma\ext A)\ra\Ty\,\Gamma \\
  & \blank[\langle a\rangle^+] && : \Ty\,(\Gamma\ext A\ext B)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]) \\
  & \blank[\langle a\rangle^{++}] && : \Ty\,(\Gamma\ext A\ext B\ext C)\ra\Ty\,(\Gamma\ext B[\langle a\rangle]\ext C[\langle a\rangle^+]) \\
  & \dots,
\end{alignat*}
and similarly $\mathbb{S}$ also includes all the (lifted) substitution
operations for terms corresponding to the top row of the
diagram. Analogously, the weakening operation corresponds to
$\blank[\p^{+\dots+}]$ operations, and the generic element is $\q$ in
our notation. The six groups of equations correspond to our
$[\langle\rangle][]$, $[\p][^+]$, $\q[^+]$, $[\p][\langle\rangle]$,
$\q[\langle\rangle]$, $[\p^+][\langle\q\rangle]$ equations, in this
order. However we don't have the lifted versions of our equations,
equations $[\langle\rangle][]$, $[\p^+][\langle\q\rangle]$ are only
stated for types, and $\q[^+]$, $\q[\langle\rangle]$ are only stated
for terms. The missing equations are admissible, see Section
\ref{sec:admissible}. In the presence of universes and $\Pi$ types, we
reduce the needed equations even more, see Section
\ref{sec:minimisation}.

@@

The first generalised algebraic presentation of type theory was Thomas
Ehrhard's calculus \cite{ehrhard,coquandEhrhard} which featured a
parallel substitution calculus with $\p$, $\blank^+$ and
$\langle\blank\rangle$ operations, just like in our calculus. Our
single substitution calculus is essentially Ehrhard's calculus with
the categorical composition and identity operations
removed. Categories with families (CwFs
\cite{DBLP:conf/types/Dybjer95,Castellan2021}) feature $\p$, $\q$,
$\blank,\blank$ operations which make it more apparent that
substitutions are lists of terms. CwFs are equivalent to Ehrhard's
calculus, and so are contextual categories/C-systems
\cite{DBLP:journals/apal/Cartmell86,DBLP:journals/lmcs/AhrensLV18},
the natural models of Awodey \cite{DBLP:journals/mscs/Awodey18} and
B-systems \cite{AHRENS_EMMENEGGER_NORTH_RIJKE_2023}.

Most computer formalisations of type theory are extrinsic
\cite{DBLP:journals/pacmpl/0001OV18,DBLP:conf/cpp/AdjedjLMPP24,DBLP:journals/jar/SozeauABCFKMTW20}:
they define the syntax as abstract syntax trees, and equip it with
typing and conversion relations. A higher level representation is
working in setoid hell \cite{chapman09eatitself}: terms are
intrinsically well-typed, but conversion is still an explicit
relation. For formalising the GAT-level syntax, one needs a stronger
metatheory than ordinary Agda or Coq: the metatheory has to support
quotient inductive-inductive types (QIITs
\cite{DBLP:journals/pacmpl/KaposiKA19}), in other words, initial
models of GATs. Altenkirch and Kaposi
\cite{DBLP:conf/popl/AltenkirchK16} formalised the syntax of type
theory using postulated QIITs in Agda, together with parametricity and
normalisation \cite{lmcs:4005} proofs. Brunerie and de Boer
\cite{initiality-agda} constructed the initial contextual category in
Agda using postulated quotients. QIITs are supported by the cubical
set model \cite{DBLP:conf/lics/CoquandHM18} and the setoid model
\cite{kaposi-qiit-setoid}, thus Cubical Agda
\cite{DBLP:journals/jfp/VezzosiMA21} and setoid/observational type
theory \cite{setoid,DBLP:phd/hal/Pujet22} support QIITs. For example,
in Cubical Agda, set-truncated and groupoid-truncated syntaxes of type
theory have been formalised \cite{cohtt}. Without QIITs, tricks such
as shallow embedding the syntax can be used to formalise metatheoretic
results about type theory such as gluing
\cite{kaposi_et_al:LIPIcs:2019:10532}, and its special cases
canonicity and parametricity \cite{kaposi-shallow}.

At the level of abstraction of SOGATs
\cite{uemura,DBLP:conf/fscd/KaposiX24}, the difference between single
and parallel substitution calculi are not visible. However, the
metatheory of type theory can be still investigated in this abstract
setting, via the methods of Synthethic Tait Computability
\cite{DBLP:phd/us/Sterling22} or internal sconing
\cite{DBLP:conf/fscd/BocquetKS23}.

@@

Notation and formalisation techniques, including congruence rules.

@@

Next we add dependent function space $\Pi$: the domain of a dependent
function is a type in some context $\Gamma$, and the codomain can also
refer to a variable in the domain, so we extend the context of the
codomain type with the type of the domain. For simplicity, both types
are at the same level (this can be remedied using $\Lift$, see later).
\begin{alignat*}{10}
& \Pi && : (A:\Ty\,\Gamma\,i)\ra\Ty\,(\Gamma\ext A)\,i\ra\Ty\,\Gamma\,i
\end{alignat*}
The nondependent function space $\blank\Ra\blank :
\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i\ra\Ty\,\Gamma\,i$ is a special case of $\Pi$
and we define it as the abbreviation $A\Ra B := \Pi\,A\,(B[\p])$.

@@

With the introduction of $\blank\cdot\blank$, we need a new
substitution rule, but again it only makes sense with an extra
equation on types saying that first substituting the last variable and
then an arbitrary substitution is the same as first the lifted version
of the arbitrary substitution that does not touch the last variable,
and then substituting the last variable.
\begin{alignat*}{10}
& [\langle\rangle][] && : B[\langle a\rangle][\gamma] = B[\gamma^+][\langle a[\gamma]\rangle] \\
& {\cdot}[] && : (t\cdot a)[\gamma] = (t[\gamma])\cdot(a[\gamma])
\end{alignat*}
Following the introduction of the operation $\langle\blank\rangle$, we
need to explain how it acts on variables. Given a variable in the
middle of the context (a term $b[\p]$), substituting its last variable
simply returns $b$. Substituting the last variable $\q$ reads out the
term from $\langle\blank\rangle$. As usual, the rules only typecheck
if we have an equation for types.
\begin{alignat*}{10}
& [\p][\langle\rangle] && : B[\p][\langle a\rangle] = B \\
& [\p][\langle\rangle] && : b[\p ][\langle a\rangle] = b \\
& \q[\langle\rangle] && : \q[\langle a\rangle] = a
\end{alignat*}
We don't have a separate sort of variables, so the rule
$[\p][\langle\rangle]$ holds not only for variables, but for arbitrary
terms. This is not an issue, as weakening a term at the end of its
context, and then substituting the newly introduced variable is the
same as not doing anything.

@@

We add the rules for universes. A universe is a type containing codes
for types, this is witnessed by $\El$ (elements) and $\cd$ (code),
which make an isomorphism between terms of type $\U\,i$ and types of
level $i$. All three operations come with substitution rules ($\El[]$
and $\cd[]$ are interderivable, so it would be enough to assume one of
the two, but we add both to completely align with the algorithm
\cite{DBLP:conf/fscd/KaposiX24} generating the rules from the SOGAT
(\ref{eq:tt})).
\begin{alignat*}{10}
  & \U && : (i:\N)\ra\Ty\,\Gamma\,(1+i)\hspace{2em} && \U[] && : (\U\,i)[\gamma] = \U\,i \\
  & \El && : \Tm\,\Gamma\,(\U\,i) \ra \Ty\,\Gamma\,i && \El[] && : (\El\,\hat{A})[\gamma] = \El\,(\hat{A}[\gamma]) \\
  & \cd && : \Ty\,\Gamma\,i\ra\Tm\,\Gamma\,(\U\,i) && \cd[] && : (\cd\,A)[\gamma] = \cd\,(A[\gamma]) \\
  & \U\beta && : \El\,(\cd\,A) = A \\
  & \U\eta && : \cd\,(\El\,\hat{A}) = \hat{A}
\end{alignat*}
Finally, we add the rules for moving types one level up. Because we
defined function space only between types at the same level, without
lifting, we cannot even define the polymorphic identity function.
\begin{alignat*}{10}
  & \Lift && : \Ty\,\Gamma\,i\ra\Ty\,\Gamma\,(1+i)\hspace{3.5em} && (\Lift\,A)[\gamma] = \Lift\,(A[\gamma]) \\
  & \mk && : \Tm\,\Gamma\,A \ra \Tm\,\Gamma\,(\Lift\,A) && (\mk\,a)[\gamma] = \mk\,(a[\gamma]) \\
  & \un && : \Tm\,\Gamma\,(\Lift\,A) \ra \Tm\,\Gamma\,A && (\un\,a)[\gamma] = \un\,(a[\gamma]) \\
  & \Lift\beta && : \un\,(\mk\,a) = a \\
  & \Lift\eta && : \mk\,(\un\,a) = a
\end{alignat*}
This concludes the first-order single substitution semantics of the
SOGAT \ref{eq:tt}. The theory can be extended with new type and term
formers in the same fashion: each rule has to be indexed over
arbitrary contexts and come equipped with a substitution rule. There
is no need to add more structural (substitution calculus) rules.

The polymorphic identity function for types at the level 0 is defined
as
\[
\lam\,(\lam\,\q) : \Tm\,\diamond\,\Big(\Pi\,(\U\,0)\,\big(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)\big)\Big).
\]
Note that readability of this term essentially relies on implicit
arguments. For example, $\q$ takes $(\diamond\ext\U\,0)$ as its first,
$1$ as its second and $\Lift\,(\El\,\q)$ as its third implicit
argument. Inrinsically typed terms are the same as derivation trees,
we illustrate this by deriving this term as follows.
\[
\infer[(*)]{\lam\,(\lam\,\q) : \Tm\,\diamond\,\big(\Pi\,(\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))\big)}{\infer{\lam\,\q : \Tm\,(\diamond\ext\U\,0)\,(\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q))}{\infer{\q : \Tm\,(\diamond\ext\U\,0\ext\Lift\,(\El\,\q))\,(\Lift\,(\El\,\q)[\p])}{}}}
\]
Note that in step (*), $\lam$ made it sure that $\U\,0$ and
$\Lift\,(\El\,\q)\Ra\Lift\,(\El\,\q)$ are types at the same level,
this is why we had to lift $\El\,\q$. Another subtle thing is
happening when deriving the third implicit argument of $\q$. We have
to coerce $\q$ along the equality $\U[]$.
\[
\infer{\Lift\,(\El\,\q) : \Ty\,(\diamond\ext\U\,0)\,1}{\infer{\El\,\q : \Ty\,(\diamond\ext\U\,0)\,0}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,(\U\,0)}{\infer{\q : \Tm\,(\diamond\ext\U\,0)\,((\U\,0)[\p])}{} && \infer{\U[] : (\U\,0)[\p] = \U\,0}{}}}}
\]
When working informally (or in extensional type theory), we don't
write the coercion, but Agda requires it.

\section{Admissible equations}
\label{sec:admissible}

TODO: say that we assume the QIIT version of this.

@@

@@

\section{Minimisation}
\label{sec:minimisation}
